---
title: "Divvy Trips 2019 User Type Analysis"
author: "Spencer Miceli"
date: "6/7/2021, updated 10/30/21"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Case Study Scenario

Lily Moreno, director of marketing and your manager, has set a clear goal: Design marketing strategies aimed at converting casual riders into annual members. In order to do that, however, the marketing analyst team needs to better understand how annual members and casual riders differ, why casual riders would buy a membership, and how digital media could affect their marketing tactics. Moreno and her team are interested in analyzing the Cyclistic historical bike trip data to identify trends.

Moreno has assigned you the question to answer: How do annual members and casual riders use Cyclistic bikes differently? 

You will produce a report with the following deliverables:
1. A clear statement of the business task
2. A description of all data sources used
3. Documentation of any cleaning or manipulation of data
4. A summary of your analysis
5. Supporting visualizations and key findings 
6. Your top three recommendations based on your analysis

## 1. Statement of the business task

The financial analysts of the company has determined that subscription members are more profitable than customer members. Analysis on how subscription members and casual riders use Cyclistic bikes differently could lead to new insights on marketing campaigns to target casual riders and convert them to subscriptions. Lily Moreno, the marketing analytics team, and the executive team could benefit from this analysis in approving and creating new marketing strategies for Cyclistic. 

## 2. Description of data sources used

The data is open public data for this case study, at [link](https://divvy-tripdata.s3.amazonaws.com/index.html). Using this data will aid in answering how Customers and Subscribers use Cyclistic bikes differently since many data points can be classified by user type.

In this analysis, data from all four quarters of 2019 will be used. The data is organized in CSV zip files per quarter of the year. The dataset likely uses a logging device for documenting when docking and unlocking at stations. Hence, the data should not be biased. Opening and filtering each dataset in Excel, we see that key metrics are not missing values either. Therefore, we know that the data is complete and accurate, and thus reliable.

The data is original since it comes from the company’s data logging system. While the data set has missing entries for birth year and gender, information such as trip_id, start_time, end_time, from_station, and to_station have no missing values.The most current Divvy_Trips data set is from 2020 Q1. Analyzing the full year of 2019 is the most current data available on the public data set. The data is used by the City of Chicago to find new ways to encourage cycling as an alternative mode of transportation to automotive vehicles. 

This is open data made available by Motivate International Inc. The data allows for analysis of how customers use bikes but does not allow for using the personally identifiable information of riders. I used sorting and filtering within Microsoft Excel to find missing values. The data helps answer my business task by seeing how subscribers and riders use Cyclistic bikes. While there are not many issues with the data, the open nature of the dataset limits the depth of analysis. For example, I cannot determine whether subscribers are more likely to live near a Cyclistic service area or if casual riders purchase multiple single passes.  

While I initially checked for missing values in Excel, I am using R to clean my data because I have over 3.8 million observations over four CSV files. Since Excel can only handle a million observations per spreadsheet, I chose R to clean my data instead since it can handle the additional entries. Using R ensures accuracy and completeness, since Excel would cut off hundreds of thousands of observations.

Additionally, weather data from the National Climatic Data Center will be used to supplement the Cyclistic data. Specifically, weather data from the O'Hare Airport in Chicago will be paired with the Cyclistic data to see how weather impacts the usage of Cyclistic bikes by Customers and Subscribers. This data can be found at https://www1.ncdc.noaa.gov/pub/data/ghcn/daily/readme.txt. The original data was pulled uploaded to a local Microsoft SQL Server database and cleaned for use with the Cyclistic data. 

The SQL query used for the weather data can be found inside the repo for this project.


## 3. Documentation of any cleaning or manipulation of data

Beginning the process of cleaning the data, load each CSV file into R:
```{r}
#Install needed packages for analysis
library(tidyverse)
library(lubridate)
library(magrittr)

#Set Working Directory to Folder with the Divvy Trips CSV Data
setwd("C:\\Users\\Spencer\\Desktop\\data\\Divvy_Trips 2019 Data")

#Load Q1-Q4 Data
Q1_df <- read_csv("Divvy_Trips_2019_Q1.csv", col_types = c("nTTnnncncccn"))
Q2_df <- read_csv("Divvy_Trips_2019_Q2.csv", col_types = c("nTTnnncncccn"))
Q3_df <- read_csv("Divvy_Trips_2019_Q3.csv", col_types = c("nTTnnncncccn"))
Q4_df <- read_csv("Divvy_Trips_2019_Q4.csv", col_types = c("nTTnnncncccn"))
```

By reading the files into R, whitespace is trimmed off automatically. I specify the column types using compact string representation so that read_csv() will give start_time and end_time the datetime data type and ensure the remaining colmuns are appropriately typed. Then, I aggregate all four quarters into one full data set for 2019. Afterwards, I remove the four quarter datasets, since we will be working with the full dataset only.

```{r}
#Bind Q1-Q4 data into a full 2019 dataset
DT_2019_df <- as_tibble(bind_rows(Q1_df, Q2_df,Q3_df, Q4_df))

#Remove extra datasets
rm("Q1_df","Q2_df","Q3_df","Q4_df")
```

I determined that for the business task at hand, Gender and Birth Year fields did not need to be analyzed. This decision was made because there are many missing or incorrect inputs from customers and these fields do not directly contribute to answering how Customers and Subscribers differ. Thus, they were removed from the dataset. Additionally, since tripduration is given in the dataset already, end_time is unnecessary and will not be of much use for analysis. Thus, the end_time will be dropped.

```{r}
DT_2019_df %<>% select(-c(gender, birthyear, end_time))
```

Lastly, I add new fields for date, month, and day of the week for further analysis. Should the analysis include more years than 2019, a column for year would be added as well. 

```{r}
#Adding start_date, start_time, month, and day of the week columns
DT_2019_df %<>% mutate(start_date = as_date(format(start_time, "%y-%m-%d"))) %>% 
               mutate(month = as.numeric(format(start_time, "%m"))) %>% 
               mutate(day_of_week = format(start_time, "%A")) %>% 
               mutate(start_time = format(start_time, "%H:%M:%S"))

str(DT_2019_df)
```

I can verify that my data is clean and ready to analyze since all of my data contains their appropriate types. Running a quick summary on the data shows that my observations are within their expected values, except for tripduration. The irregularities in tripduration will be explored further in the analysis section.


## 4. In-depth Analysis and Supporting Visualizations

First, we see that of all rides in 2019, 2.93 million (77%) of rides were completed by subscribers and 880,000 (23%) were completed by customers.
```{r}
table((DT_2019_df$usertype))
prop.table(table((DT_2019_df$usertype)))
```

Checking summary statistics of Subscribers and Customers, we see that while the maximum and minimum trip duration for both users are similar, their mean and median trip durations differ.
```{r}
#Compare trip duration summary statistics between customers and subscribers
DT_2019_df %>% select(tripduration, usertype) %>% group_by(usertype) %>%  
    summarize(min=min(tripduration),
              Q1 = quantile(tripduration, 1/4),
              mean=mean(tripduration), 
              median = median(tripduration),
              Q3 = quantile(tripduration, 3/4),
              max=max(tripduration), 
              IQR=IQR(tripduration),
              predicted_maximum = quantile(tripduration, 3/4) + 1.5*IQR,
              num_of_outliers = sum(tripduration > predicted_maximum),
              std_dev = sd(tripduration),
    )
```

The accompanying boxplot for the above summary statistics: 

```{r}
DT_2019_df %>% group_by(usertype) %>% ggplot() + 
  geom_boxplot(mapping = aes(x=tripduration, group = usertype, fill = usertype)) +
  coord_cartesian(xlim = c(0,6000)) +
  theme(axis.ticks.y = element_blank(),
        axis.text.y = element_blank()
  ) +
  labs(title = "Boxplot for Trip Duration by User Type")
  
```

The mean trip duration is much larger than the median trip duration for both Customers and Subscribers. This is due to the large number of outliers for both user types that skew the means. Thus, we will look mostly at the median trip duration for this analysis.

Next, we will look at the median trip duration for Subscribers and Customers by day of the week.

```{r}
#Correctly order the days of the week, then check the mean trip duration by day and user type
DT_2019_df %>% group_by(usertype, day_of_week) %>%  summarize(mean=mean(tripduration), median = median(tripduration),
                                                 max=max(tripduration), min=min(tripduration), IQR=IQR(tripduration),
                                                 std_dev = sd(tripduration) )
```

To supplement the data, we'll construct two visualizations:
```{r, echo=FALSE}
#Visualize median trip duration by day of the week and user type
DT_2019_df %>% 
  group_by(usertype, day_of_week) %>%                     #group by usertype and weekday
  summarise(number_of_rides = n(), average_duration = mean(tripduration),
            median = median(tripduration) ) %>%       #calcuates mean and median tripduration
  arrange(usertype) %>%                               #sorts by usertype
  ggplot(aes(x=day_of_week, y = median, fill = usertype)) +
    geom_col(position = "dodge")+                     #plots a barchart for median
    labs(x = "Day of the Week",
         y = "Median Trip Duration",
         title = "Median Trip Duration by Day of the Week and User Type in 2019")

```

We see two general trends emerge from these visualizations. First, Customers take longer trips than subscribers. Secondly, both Subscribers and Customers ride for slightly longer durations on the weekends.

Next, I analyze which stations our customers starting their trips from. To do this, first I create a new table that shows corresponds the station ID with the number of rides started at that station. Then, I run some summary statistics on these tables.
```{r}
#Creating tables for number of Trips based on Subscribers and Customers
Sub_DT_df <- DT_2019_df %>% filter(usertype == "Subscriber") %>% select(from_station_id) %>% 
  group_by(from_station_id)%>% summarize(count = n()) 

Cus_DT_df <- DT_2019_df %>% filter(usertype == "Customer") %>% select(from_station_id) %>% 
  group_by(from_station_id)%>% summarize(count = n())

#Summary Statistics
(Sub_DT_ss <- Sub_DT_df %>% summarize(
                min=min(count),
                Q1 = quantile(count, 1/4),
                mean=mean(count), 
                median = median(count),
                Q3 = quantile(count, 3/4),
                max=max(count), 
                IQR=IQR(count),
                predicted_maximum = quantile(count, 3/4) + 1.5*IQR,
                num_of_outliers = sum(count > predicted_maximum),
                std_dev = sd(count)
              )
)
(Cus_DT_ss <- Cus_DT_df %>% summarize(
                min=min(count),
                Q1 = quantile(count, 1/4),
                mean=mean(count), 
                median = median(count),
                Q3 = quantile(count, 3/4),
                max=max(count), 
                IQR=IQR(count),
                predicted_maximum = quantile(count, 3/4) + 1.5*IQR,
                num_of_outliers = sum(count > predicted_maximum),
                std_dev = sd(count)
              )
)
```

We see that our Subscribers do not seem to start from a group of stations more than others. However, for customers, the mean is close to the 3rd. Quartile, indicating that there are stations that are receiving a lot of traffic for customers. 

For this report, a "high volume station" will be any station that has an annual number of start rides that is greater than the predicted maximum, which is calculated using the formula: 

Predicted maximum = Q3 + 1.5*(Q3-Q1).

Using the quartile statistics given, and the predicted maximum formula, we can find if the percentage of high volume stations that subscribers and customers are starting at much more frequently than other stations.

```{r}
tibble(
  Subscriber = paste(round(as.numeric(Sub_DT_df %>% 
                                        filter(count > Sub_DT_ss$predicted_maximum) %>%
                                        summarize(total = sum(count))/as.numeric(Sub_DT_df %>% 
                                                      summarize(total= sum(count)))
                                      )*100, 2), "%")
,


  Customer = paste(round(as.numeric(Cus_DT_df %>% 
                                        filter(count > Cus_DT_ss$predicted_maximum) %>%
                                        summarize(total = sum(count))/as.numeric(Cus_DT_df %>% 
                                                      summarize(total= sum(count)))
                                      )*100, 2), "%")
)
```

We see that 49% all Customers start their rides from one of 49 high volume station, whereas only 21.5% of Subscribers start from one of 25 high volume station. Let's create an accompanying visual to illustrate.

```{r, echo=FALSE}

#Bar Chart for High Volume Stations for Customers in 2019

Cus_DT_df %>% 
  mutate(rownum = row_number()) %>% arrange(desc(count)) %>% slice(1:25) %>% 
  ggplot(aes(x=reorder(rownum, count), y = count, fill = from_station_id,))+
    geom_col(color = "black")+
    labs(title = "Percentage of Rides at the Top 25 Highest Volume Stations for Customers in 2019",
         subtitle = "All stations with more than 5300 rides annually",
         x = "Station ID",
         y = "Percentage of Rides") + 
    guides(fill = "none")+
    theme(axis.ticks.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.y = element_blank(),
          axis.text.y = element_blank(),
          panel.grid = element_blank(), 
          panel.background = element_blank(),
          axis.title = element_text(size = 10, face = "bold"),
          title = element_text(size = 10, face = "bold"))+
    geom_text(aes(label = (paste(from_station_id)),x = reorder(rownum, count), y = -0.3),
              vjust = -0.5, 
              size = 3)+
    geom_text(aes(label = (paste(round(count, 1))),x = reorder(rownum, count)),
           vjust = -1,
           color = "black", size = 2.5)+
    scale_fill_gradient(low = "yellow1", high = "cyan1")
```

```{r}
tibble(
  Percentage_of_rides = paste(round(as.numeric(
                              Cus_DT_df %>% arrange(desc(count)) %>% slice(1:25) %>% 
                              summarize(total_rides = sum(count)) / as.numeric(Cus_DT_df %>% 
                                                      summarize(total= sum(count)))
                              )*100, 2), "%")
  ,

  Percentage_of_stations = paste(round(as.numeric(
                              Cus_DT_ss$num_of_outliers / 
                              Cus_DT_df %>% select(from_station_id) %>% summarize(n = n())
                              )*100, 2), "%")
)
  #percentage of customer rides started at one of the top 25 highest volume stations
```

Since displaying all 49 stations in the bar chart would be quite cluttered, the Top 25 stations for Customers are shown.

We see from this visual the station ID of the top 25 high volume stations for customer rides for 2019. Recall that  49 stations account for 49% of all rides from customers in 2019. These top 25 high volume stations are significant since these stations account for 8% of all our stations but yet accrue 37.66% of all rides for customers. Many of these stations could be prime locations for increased and targeted marketing.

```{r}
print( Cus_DT_df %>% arrange(desc(count)) %>% slice(1:25) %>% left_join(DT_2019_df, by = "from_station_id") %>%
        distinct(from_station_name)
  
)
```


We could display a similar visual for Subscribers, but 25 high volume stations for subscribers make up 21.5% of the total number of rides. This is not as significant as Customers, which indicates that Subscribers make greater use of more stations than Customers. 


Lastly, we will visualize Customers and Subscribers median trip duration based on the month of the year.

```{r, echo=FALSE}
#Bar Chart of Customers and Subscribers Median Trip Duration by Month
DT_2019_df %>% 
  group_by(usertype, month) %>%                    #group by usertype and month
  summarise(number_of_rides = n(), average_duration = mean(tripduration), 
            median = median(tripduration) ) %>%      #calcuates mean and median tripduration
  arrange(usertype) %>%                              #sorts by usertype
  ggplot(aes(x=month, y = median, fill = usertype)) +
    geom_col(position = "dodge")+                     #plots a barchart for mean
    labs(x = "Month",
       y = "Median Trip Duration",
       title = "Median Trip Duration by Month and User Type in 2019")
```

We see that Customers ride longer trips during the spring and summer months than in the fall and winter months. Subscribers follow a similar pattern, with much less variability. Next, we will see how total number of rides changes by month.

```{r, echo=FALSE}
#Bar Chart of Customers and Subscribers Total Number of Trips by Month
DT_2019_df %>% 
  group_by(usertype, month) %>%                    #group by usertype and month
  summarise(number_of_rides = n(), average_duration = mean(tripduration), 
            median = median(tripduration) ) %>%      #calcuates mean and median tripduration
  arrange(usertype) %>%                              #sorts by usertype
  ggplot(aes(x=month, y = number_of_rides, fill = usertype)) +
  geom_col(position = "dodge")+                     #plots a barchart for mean
  labs(x = "Month",
     y = "Number of Trips",
     title = "Number of Trips by Month and User Type in 2019")

```

This visual shows us that Subscribers and Customers take more trips during the spring and summer months than in the fall and winter months. In fact, Customers take very few trips from November through February, whereas Subscribers still see much use of the bicycles during the winter months. Lastly, we will visualize how use of the bike share changes by time of day.


```{r}
DT_2019_df %>% select(start_time, usertype) %>%  mutate(start_hour = hour(hms::as_hms(start_time))) %>% 
  filter(usertype == "Customer") %>% ggplot() +
  geom_histogram(mapping = aes(x=start_hour), color = "black", fill = "darkcyan", bins = 24)+
  labs(title = "Number of Rides for Customers by Hour of Day",
       x = "Hour",
       y = "Number of Trips")
DT_2019_df %>% select(start_time, usertype) %>%  mutate(start_hour = hour(hms::as_hms(start_time))) %>% 
  filter(usertype == "Subscriber") %>% ggplot() +
  geom_histogram(mapping = aes(x=start_hour), color = "black", fill = "coral", bins = 24)+
  labs(title = "Number of Rides for Subscribers by Hour of Day",
       x = "Hour",
       y = "Number of Trips")

```

We see that customers typically take rides between noon and early evening. On the contrary, subscribers take rides mostly during the morning and evening rush hours. This indicates that customers are taking rides for different reasons than subscribers. Subscribers use Cyclistic for commute, whereas customers use it for leisure. 

We saw that Customers are less likely to take any rides during the winter months. Since customers seem to ride primarily for leisure, then weather likely impacts how customers use the bikes. Now, we'll add an additional data set from the National Climatic Data Center, taking weather data from the O'Hare airport in Chicago. 

```{r}
weather <- as_tibble(read_csv("OHare_Airport_Weather_2019.csv", col_names = TRUE))
weather_DT_df <-DT_2019_df %>% 
  select(start_date, tripduration, usertype) %>% 
  left_join(weather, by = c("start_date" = "date"))
```




```{r}
bind_rows(
weather_DT_df %>% filter(PRCP < 254) %>% group_by(usertype) %>% 
  summarize(condition = "Precipitation < 1 inch",
            number_of_trips = n(), 
            avg_number_of_trips = n() / as.numeric(
              weather_DT_df %>% filter(PRCP < 254) %>% distinct(start_date) %>%
              summarize(number_of_days = n())
              ),        
            median = median(tripduration),
            std = sd(tripduration)
            )
,
weather_DT_df %>% filter(PRCP >= 254) %>% group_by(usertype) %>% 
  summarize(condition = "Precipitation >= 1 inch",
            number_of_trips = n(),
            avg_number_of_trips = n() / as.numeric( 
              weather_DT_df %>% filter(PRCP > 254) %>% distinct(start_date) %>%
              summarize(number_of_days = n())
              ),
            median = median(tripduration), 
            std = sd(tripduration)
            )
,

weather_DT_df %>% filter(SNOW < 12.7) %>% group_by(usertype) %>% 
  summarize(condition = "Snow < 0.5 inch",
            number_of_trips = n(), 
            avg_number_of_trips = n() / as.numeric(
              weather_DT_df %>% filter(SNOW < 12.7) %>% distinct(start_date) %>%
              summarize(number_of_days = n())
              ),
            median = median(tripduration), 
            std = sd(tripduration)
            )
,
weather_DT_df %>% filter(SNOW >= 12.7) %>% group_by(usertype) %>% 
  summarize(condition = "Snow >= 0.5 inch",
            number_of_trips = n(), 
            avg_number_of_trips = n() / as.numeric(
              weather_DT_df %>% filter(SNOW > 12.7) %>% distinct(start_date) %>%
              summarize(number_of_days = n())
              ),
            median = median(tripduration), 
            std = sd(tripduration)
            )
,
weather_DT_df %>% filter(AWND < 67.06) %>% group_by(usertype) %>% 
  summarize(condition = "Average Wind Speed < 15mph",
            number_of_trips = n(), 
            avg_number_of_trips = n() / as.numeric(
              weather_DT_df %>% filter(AWND <= 67.06) %>% distinct(start_date) %>%
              summarize(number_of_days = n())
              ),
            median = median(tripduration), 
            std = sd(tripduration)
            )
,
weather_DT_df %>% filter(AWND > 67.06) %>% group_by(usertype) %>% 
  summarize(condition = "Average Wind Speed >= 15mph",
            number_of_trips = n(), 
            avg_number_of_trips = n() / as.numeric(
              weather_DT_df %>% filter(AWND > 67.06) %>% distinct(start_date) %>%
              summarize(number_of_days = n())
              ),
            median = median(tripduration), 
            std = sd(tripduration)
            )
)
```

Here, we see that the weather conditions change how customers use Cyclistic bikes. We see that with adverse weather conditions, Customers tend to not use Cyclistic bikes. Windy and Snowy days are prime examples of how customers use the bikes less frequently. While subscribers also see a decrease in number of trips, it isnt as large as the decrease for customers. 

Lastly, the number of bike rides for customers is plotted against the maximum recorded temperature to see if there is a relationship between good weather and Customer rides.

```{r}
weather_DT_df %>% select(TMAX, usertype, tripduration) %>% mutate(temp_F = (TMAX/10)*(9/5)+32) %>% 
  filter(usertype == "Customer") %>% 
  group_by(temp_F) %>% summarize(number_of_rides = n()) %>% 
  ggplot() + geom_point(mapping = aes(x=temp_F,y=number_of_rides)) +
  labs(title = "Temperature in Fahrenheit vs Number of Customer Rides",
       x = "Temperature in Fahrenheit",
       y = "Number of Rides")
```

We see that the number of customer rides do increase with good weather. There could be other factors at play that determine the increase in ridership though, such as more tourists riding bikes in Chicago during months with better weather.

## 5. Summary of Analysis

The most important findings from this analysis are the following.

* Customers have an average and median trip duration that is longer than the average and median trip duration of Subscribers.

* 49% of all Customer rides started at 49 high volume stations for Customers.

* Both trip duration and number of rides are higher among both Subscribers and Customers during the spring and summer months.

* Customers make far less use of the bike share from November through February.

* Adverse weather conditions greatly discourage Customers from using Cyclistic bikes.


## 6. Recommendations
Based on the summary of the analysis section, there are three recommendations that I have for Cyclistic.

1. Develop new Subscriber pricing models to appeal to Customers who use Cyclistic bikes for leisure. 

2. Investigating high volume stations. 49% of customers start at 49 stations. Determining why customers tend to use these stations more than others could result in developing stronger marketing strategies.

3. Develop seasonal marketing strategies. Marketing to customers in the spring and summer could result in more subscriptions since there are more customers wanting to ride on the bike share. Additionally, marketing discounts during the fall and winter months could result in increased ridership during the fall and winter. 
